---
title: 'Lab 8: Analysis of Variance'
author: "NAME"
format: html
embed-resources: true
html-math-method: mathml
execute: 
  warning: false
  echo: true
  error: false
---


## Before you start:

1. Complete the Lab Primer linked in Canvas.  
2. Be sure you have opened the `lab8-anova-student.Rproj` file.  
3. Change the name of the **QMD** file only to your last name, first initial, e.g., `OtterM-lab8-anova-student.qmd`. You will have to close the QMD document to rename the file.   
4. Change the "NAME" above after `author:` to your name, leaving it in quotes: "Monte Otter"  


## Skills

*	Explain the purpose of a multiple comparison procedure to compare multiple means.
* Be able to identify a situation that requires ANOVA.  
    + Identify the factor (i.e. explanatory variable) and response variable.  
    + Be able to state the appropriate hypotheses for ANOVA.  
    + Report ANOVA results in an ANOVA table and be able to interpret each part of the table within the context of the study.  
* Explain the necessary conditions to use an ANOVA  
    + Be able to assess the conditions.  
* Be able to identify a situation that requires an ANOVA for transformed data
    + Identify the factor (i.e. explanatory variable) and response variable
    + Be able to state the appropriate verbal-only hypotheses for a transformed ANOVA
    + Report ANOVA results in an ANOVA table and be able to interpret each part of the table within the context of the study.
* Be able to interpret output from a post hoc test (Tukey's method).
* Be able to explain the difference between multiple t-tests and a post hoc test.
* Be able to execute the proper R codes to conduct ANOVA and post hoc analyses.

The above skills cover LT 1-5, 10-16, 21c, 22ab, and 23a.  It also introduces R-LT R7. 


Here are some symbols that you might find useful:  
$H_0$ $H_1$ $\sigma$ $\mu$ $\mu_0$ $\mu_{word}$ $\mu_{d~word}$ $\neq$


Update the following code chunk with the appropriate packages.  

```{r}
#| label: setup
#| include: false
library(tidyverse)

```



***

## Scenario 1: Tree Moisture    

Researchers from the Department of Botany at Duke University were interested in studying the moisture content of trees. Each of the following four species were sampled: loblolly pine (*Pinus taeda L.*), shortleaf pine (*Pinus echinata Mill.*), yellow poplar (*Liriodendron tulipifera L.*), and red gum (*Liquidambarstyracifua L.*). Sampling consisted of collecting a single branch from a tree to measure its moisture content. The researchers randomly chose to cut each branch in one of three locations (proximal, central, or distal, each further along the branch from the trunk). 

All trees sampled were about ten to twelve feet high and were growing in open stands where the lower limbs were well exposed to light. The branches chosen for sampling were about four feet above the ground, and only one branch was sampled from any one tree. The series of tests was made between 11 a.m. and 1 p.m. on a cool, moist, cloudy day in July when transpiration (the loss of water through the leaves to the atmosphere) was presumably slow for all trees. As the samples were cut, they were placed in closed weighing cans and taken into the laboratory where the wet weights were obtained. They were dried to constant weight in a forced draft oven at 800 degrees C and reweighed, the loss in weight being taken in each instance to represent the water content of the fresh sample. The moisture content, in percent, is expressed as a percentage of the weight of the water as compared to its oven dried weight. The data are found in `tree-moist.csv`.

**Researchers want to determine if there is a difference in the moisture content between the four different tree species.**


<Insert code chunk here and read in the data> 



### 1.1. 
Since the researchers plan to make comparisons among the four species of trees, they chose first to conduct an ANOVA test on the tree moisture content based on species before making comparisons between each species of tree. Why is it better to conduct an ANOVA first before making comparisons between each pair of tree species?

> Insert answer here   


### 1.2. 
Name the response and explanatory variables of interest and, for each, state the type of variable (quantitative or qualitative; continuous, discrete, nominal, or ordinal). Recall this research question is examining the effect of species on moisture content of the tree branches.

> Response: Insert answer here    
> Explanatory: Insert answer here   


### 1.3. 
State the null and alternative hypotheses, symbolically and verbally as appropriate.

> Insert answer here   


### 1.4. 
Modify the code below to conduct an appropriate testing of the conditions to calculate an ANOVA analysis. State each condition, state the evidence you used to assess the condition (e.g. graphic), and then assess the evidence for if the condition is met or not. Use the template below the code chunk for your answer to each condition.  

```{r}
#| label: condition-checks

species_aov <- [insert code that does an anova here]

<insert code that creates a qq plot of residuals here>

<insert code that creates a residuals plot here>
```

> Condition #_: ___     
> Evidence: ___    
> Justification: ___    

> insert more condition templates as needed  



### 1.5. 
Insert a code chunk and write the code below to print out an anova table for your analysis. Fill in the values of the appropriate statistics.

<Insert code chunk here>   

> df~e~:    
> df~t~:    
> MSE:    
> MST:     
> F:    
> p-value:   


### 1.6. 
Interpret your p-value in the context of the question.

> Insert answer here   


### 1.7. 
Write a strength of evidence statement, using the significance level $\alpha = 0.025$  

> Insert answer here   


### 1.8. 
Based on your exploratory data analysis (EDA), what would be your conclusion in answer to the research question about if there are differences in moisture content among the tree species? You will need to insert a code chunk to calculate the appropriate statistics and create a graphic to answer this question. Be sure to cite your evidence and justify how it supports the conclusions about the parameters.   

```{r}
#| label: exploratory-data-analysis
<insert summary statistics code here>
<insert boxplot code here>
```

> Insert answer here   



### 1.9. 
Is there any reason to do a Tukey test on the moisture content? **Bold** the words (but not the list-letter) of your answer.   

a. No, we risk elevated Type I error with multiple comparisons in a Tukey test.  
b. No, there is no reason to need a comparison.  
c. Yes, we should do a Tukey test to test our original null hypothesis.   
d. Yes, we should do a Tukey test to identify which pairs are significantly different.  


### 1.10. 
What is the consequence of using an independent two-sample t-test for each pairwise comparison (instead of using the Tukey test)? **Bold** the words (but not the list-letter) of your answer.  

a. We risk an increased Type I error rate in each test.  
b. We risk an increased chance of committing a Type I error.  
c. We risk an increased chance of committing a Type II error.  
d. We risk eliminating random coincidences as a reasonable explanation.  


### 1.11. 
Conduct a Tukey test by modifying the code below. (1) Fill out the table below to organize the results and (2) then provide a written summary of your *post hoc* test results of all possible factor pairings. In your written summary, you should address all pairwise comparisons in terms of if they are the same, if they are different (i.e., at least moderately strong evidence against the null, using the family-wise significance of 0.05, and include p-values as evidence), and, if it exists, the direction of difference. *This is not a list. Simply state generally how parameters compare, the way you might communicate verbally with a scientist, with evidence.*

```{r}
#| label: tukey-comparisons
______(species_aov)
```


  pair   | p-value | strong evidence? (Y/N) | how do they compare? (use <,>,or =)
---------|---------|------------------------|---------------------------------------
  RG-LP  |         |                        |    
  SP-LP  |         |                        |    
  YP-LP  |         |                        |  
  SP-RG  |         |                        |    
  YP-RG  |         |                        |    
  YP-SP  |         |                        |    


> Insert answer here  


 
### 1.12. 

Woodworkers need their lumber to have a moisture content of around 8-12%. The acceptable moisture levels of wood for exterior wood are at the higher end of this range, while interior wood projects require the lowest moisture levels. A woodworker wants to build indoor cabinets for his kitchen. What would you advise based on the results of the study above?    

> Insert answer here   


### 1.13. 
Do you trust the results of your ANOVA and Tukey tests? Explain why with statistical reasoning. If you do not trust the results, suggest and justify an alternate analysis. 

> Insert answer here   


### 1.14. 

Given our scenario, **bold** any answers (but not the list-letter) below that is a true statement.  

a. The tree branches were sampled randomly, so this can only be an observational study and thus we cannot determine causality.  
b. The tree branches were sampled randomly, so this can only be an observational study and thus we can generalize to all the tree branches.  
c. The tree branches were not randomly assigned to each treatment group (species), so this is an observational study and thus we can generalize to all the tree branches.  
d. The tree branches were not randomly assigned to each treatment group (species), so this is an observational study and thus we cannot determine causality.  




***



### Scenario 2: Ant Foraging     


Red wood ants, *Formica rufa*, forage for food (mainly insects and ‘honeydew’ produced by aphids) both on the ground and in the canopies of trees. Rowan, oak and sycamore trees support very different communities of insect herbivores (including aphids) and it would be interesting to know whether the foraging efficiency of ant colonies is affected by the type of trees available to them. As part of an investigation of the foraging of *Formica rufa*, observations were made of the prey being carried by ants down trunks of randomly sampled rowan, oak and sycamore trees. The total biomass of prey being transported was measured over a 30 minute sampling period and the data were expressed as the biomass (dry weight in mg) of prey divided by the total number of ants leaving the tree to give the rate of food collection per ant per half hour. The data are in `ant-foraging.csv`.    

**Researchers want to determine if the total biomass of prey being transported by *Formica rufa* is different across the three tree species types**. 


### 2.1. 
State the null and alternative hypotheses, symbolically and verbally as appropriate.

> Insert answer here   


### 2.2. 
State each condition, state the evidence you used to assess the condition, and then assess the evidence for if the condition is met or not. Be sure to fix your explanatory variable for an ANOVA analysis.     


<Insert code chunk here>   

> Insert answer here   



*** 
If the parametric conditions of ANOVA are not met, we can consider using a transformation before using a non-parametric option. In your class notes, we gave you some general guidelines for choosing possible transformations to try on your data. We will perform a square root transformation on the data from each tree species using the `sqrt()` function. 


### 2.3. 
*Challenge Question* What do you notice about the distribution of red wood ant rate of food collection that might suggest a square root transformation would be appropriate? *(Hint: a boxplot, calculating summary statistics, or looking back at your condition plots might help you determine this)*

<Insert code chunk here>  

> Insert answer here   

 

### 2.4. 
Name the explanatory and the response variables of interest for the **square-root-transformed** ANOVA and state the type of variable (quantitative or qualitative; continuous, discrete, nominal, or ordinal). 

> Response: Insert answer here    
> Explanatory: Insert answer here     



### 2.5. 
State the null and alternative hypotheses for the **square-root-transformed** ANOVA, verbally only.

> Insert answer here   


### 2.6. 
Assess all the conditions of the **square-root-transformed** ANOVA. State each condition to conduct the inference test, state the evidence you used to assess the condition, and then assess the evidence for if the condition is met or not.    


```{r}
#| label: sqrt-aov

sqrt_ant_aov <- <insert function>(sqrt(<insert reponse>) ~ factor(<insert explanatory>), data = <insert dataname>)


```


> Insert answer here   



### 2.7. 
Insert a code chunk that will print out your ANOVA table from the *square-root-transformed* analysis. Fill in the appropriate statistics below. 

<Insert code chunk here>   


> df~e~:     
> df~t~:     
> MSE:     
> MST:      
> F:     
> p-value:     


### 2.8. 
Write a statement of the strength of your evidence from the **square-root-transformed** ANOVA.   

> Insert answer here   


### 2.9. 
If the ANOVA on transformed data yields significant differences between the three tree species, we will want to conduct a *post-hoc* analysis. Have we met the conditions to perform a post-hoc Tukey test? (yes or no & why; do not re-assess).

> Insert answer here   


### 2.10. 
Insert a code chunk to conduct a Tukey test. Provide a written summary of your *post hoc* test results of all possible site pairings. 

< Insert code chunk here>    

> Insert answer here   



### 2.11. 
Given our scenario, **bold** any answers (but not the list-letter) below that is a true statement.

a. The trees were not randomly assigned to each treatment group, so this is an observational study.    
b. The trees were sampled randomly, so this can only be an observational study and thus we cannot determine causality.   
c. The trees were not sampled randomly, so we cannot generalize to all locations within the three species.   
d. The trees were sampled randomly, so we can generalize our results to all trees at these sites.   
e. The trees were randomly assigned to each treatment group, so this is an experiment, so we can say any changes we observed is due to the tree species.   
f. The trees were randomly assigned to each treatment group, so this is an experiment, but we cannot determine the cause of the difference.  




