---
title: 'Lab 5: Two Group Comparison EDA & QQ Plots'
author: "NAME"
format: html
embed-resources: true
html-math-method: mathml
execute: 
  warning: false
  echo: true
  error: false
---


## Skills

* Identify the explanatory and response variables in an experiment or observational study  
* Identify the type of study and its explanatory and response variables  
* Identify types of randomization in a study  
* Be able to produce appropriate graphics and summary statistics for independent two-sample and paired data  
* Be able to interpret data from a boxplot and histogram  
* Understand how a QQ plot is created and why it can assess normality
* Be able to assess normality of a sample based on the QQ plot  

The skills in this Lab include LT 11-13, 22ab, in addition to R1 and R2.


Here are some symbols that you might find useful:  

$\sigma$ 

$\mu$ 

$\mu_{word}$ 

$\mu_{d~word}$


***

Before you start the lab, add the code that will load any other packages you might need into this code chunk. 

```{r}
#| label: setup
#package for data management
#package for graphics
#package for stats

```



## Scenario 1: Fat Bears

Katmai National Park is famous for the salmon run each summer and the brown bear feast that accompanies it. Brown bears can lose roughly 30% of their body weight while hibernating, so it is essential that they pack on the pounds during the summer. They feed on high-caloric foods, like sockeye salmon (one salmon can be ~4000 calories), but also on sedges (a grass, often eating about 30 lbs of sedges a day), berries, grubs, etc. Over the course of the summer, bears have to add hundreds of pounds to their body weight to make up for weight lost during the past hibernation, and to survive the next hibernation; at peak season, brown bears can add up to four pounds of weight per day! Over the summer, the brown bears can really pack it on! (case in point: see pictures of the [Katmai Fat Bear champion final contenders](https://explore.org/fat-bear-week)

You are a researcher interested in the weight gain of brown bears over the summer. Each bear in the park is tagged with a unique identification number. You weigh a random selection of brown bears in Katmai NP at the beginning of June, and then weigh the same bears at the end of September and record their weight (in pounds). 

This code reads in our data and then prints a list of the vector names to the screen.

```{r}
#| label: load-bears
bears <- read_csv("fat-bear-weights.csv", show_col_types = FALSE)
names(bears)
```


### 1.1. 
Identify each variable of interest in our analysis and the type of variable for each by replacing the [placeholders] below.

Variable        | Explanatory or Response Variable?  | Numeric or Categorical?
----------------|------------------------------------|-------------------------
[placeholder]   | [placeholder]                      | [placeholder]           
[placeholder]   | [placeholder]                      | [placeholder]           



### 1.2. 
State the parameter(s) of interest in the study.

> Insert answer here  


### 1.3. 
State the study design of this study (two independent samples or one matched-pair sample). Justify your response.

> Insert answer here   
> Insert justification here  


### 1.4. 
State the study type of this study (observational or experiment?). Justify your response. 

> Insert answer here  
> Insert justification here  


### 1.5. 
Add code below to calculate the summary statistics (listed below) for *differences* in the weight gained over the summer. Place your resulting values in the table below, rounding to four decimal places as needed. 

```{r}
#| label: stats-bears
(gain_n <- length(bears$sept_lb)) #notice we just have to include one variable for n
(gain_xbar <- mean(~(____ - _____), data = ____)) #calculate mean difference
(gain_var <- _________________________) #calculate the variance of the difference
(gain_s <- ___________________________) #calculate the sd of the differences
```

Symbols     | Weight~diff~ 
------------|---------------
$n$         | [placeholder]              
$\bar{x}_D$ | [placeholder]              
$s_D^2$     | [placeholder]              
$s_D$       | [placeholder]   


### 1.6. 
Create a boxplot for the weight gained over the summer. Be sure to label the axes appropriately. 

< Insert code chunk here >  


### 1.7. 
Describe the distribution of weight gained over summer in brown bears based on your boxplot (e.g., center, spread, shape, outliers). 

< Insert code chunk here >  

> Insert answer here  


### 1.8 
Produce a qqplot for the differences between September and June weights for the bears.

```{r}
#| label: qq-bears
gf_qq(~(_______ - ______), data = _________) |> 
  gf_qqline()
```


### 1.9 
Assess the plot for normality, including a justification.

> Insert answer here   


Want more Fat Bear before and after pictures? Check out the [this page](https://explore.org/meet-the-bears) from this fall's Fat Bear Competition of all the contenders!  

Note -- it would be *really* hard to actually weigh a grizzly bear; researchers at Katmai are working to develop new tools to measure bear volume with 3D laser imaging: [lidar in 2020](https://www.popsci.com/story/technology/fat-bear-week-laser-scanning-weight-estimates/) and [volume to weight ratios from 2019](https://medium.com/@exploreorg/how-big-are-fat-bear-week-bears-cd4f23036c33).



 
## Scenario 2: Exxon Valdex Oil Spill   

Following the March 24, 1989 grounding of the oil tanker the Exxon Valdez in Alaska, approximately 35,500 tons of crude oil was released into Prince William Sound.  The paper, “The deep benthos of Prince William Sound, Alaska, 16 months after the Exxon Valdez oil spill” [Marine Pollution Bulletin (1998), 36: 118-130] reports on an evaluation of the organisms, benthic macrofauna, that live in the sea-bed after the spill.  Thirteen sites were selected for study.  Seven of the sites were within the oil spill trajectory (affected by the oil) and six were outside the oil trajectory (not affected by the oil).  Collection of the data occurred at two depths, 40m and 100m, in the period of July 1-23, 1990.  One of the variables measured was abundance of the benthic macrofauna (individuals per square meter) at both 40m and 100m.  The measures of abundance at each depth are represented in our data as `Depth40m` and `Depth100m`.  **For our study, we will be focusing on the abundance of the benthic macrofauna (individuals per square meter) at 40m to determine if there is still an effect of the oil spill by comparing site locations within the oil spill trajectory and outside the oil spill trajectory.**  


### 2.1. 
Insert a code chunk that will read in your csv file `oil-spill.csv`, save it as the object `oilspill`, and then print a list of the vector names to the screen. (*Note: You can __NOT__ use `View`*.)

< Insert code chunk here >  


### 2.2. 
Identify each variable of interest in our analysis and the type of variable for each by replacing the [placeholders] below.

Variable           | Explanatory or Response Variable?  | Numeric or Categorical?
-------------------|------------------------------------|-------------------------
[placeholder]      | [placeholder]                      | [placeholder]           
[placeholder]      | [placeholder]                      | [placeholder]           



### 2.3. 
State the parameter(s) of interest in the study.

> Insert answer here  


### 2.4. 
State the study design of this study (two independent samples or one matched-pair sample?). Justify your response.

> Insert answer here   
> Insert justification here  


### 2.5. 
State the study type of this study (observational or experiment?). Justify your response. 

> Insert answer here    
> Insert justification here  


### 2.6. 
Calculate the relevant summary statistics. Annotate your code using `#` to describe what each line of code is doing. Enter your values in the table below.

< Insert code chunk here >  


Symbols   |   Outisde     |  Within        
----------|---------------|---------------
$n$       | [placeholder] | [placeholder]              
$\bar{x}$ | [placeholder] | [placeholder] 
$s^2$     | [placeholder] | [placeholder] 
$s$       | [placeholder] | [placeholder] 


### 2.7. 
Create a boxplot of each group in your sample in one single graphic (i.e., numerical variable by categorical variable). Be sure to label the axes appropriately.

< Insert code chunk here >  


### 2.8. 
Without conducting the hypothesis test, what might you conclude about the abundance of the benthic macrofauna (individuals per square meter) at 40m between sites **outside** the oil spill trajectory and sites **within** the oil spill trajectory, based solely on the provided exploratory data analysis in the previous questions? Justify your conclusion about the parameter(s) using the graphics and summary statistics. 

> Insert answer here      


### 2.9. 
Produce a qq plot for each of the abundance of the benthic macrofauna at 40m for sites by location (outside or within the oil spill trajectory). 

< Insert code chunk here >   


### 2.10. 
Assess each plot for normality, including a justification. 

> Insert answer here     

