---
title: "Lab 1: Introduction to R"
author: "KEY"
format: html
embed-resources: true
html-math-method: mathml
execute: 
  warning: false
  echo: true
  error: false
---




<span style='color:#000000;background:#ffc0cb;'>
- Use only the following snippets for this lab: cc, ok, fb, edit, render, com, cor  
- Comments only for naming issues:  
*Naming system should be LastF_lab.... but as long as sorts correctly in your file folder, it is fine.*</span>

## Skills

* Demonstrate basic skills in R coding
* Recognize R syntax to decode new R commands
* Be able to identify basic data types in R and how to encode them (see Primer)
* Be able to create and manipulate vectors in R
* Be able to create and manipulate dataframes in R
* Be able to use variables in arithmetic functions in R

## Learning Targets

Learning Targets will be identified prior to questions as appropriate. Not every question will directly align with a LT -- those questions are present to help you develop into the skills needed for the LTs; labs are expected to be fully completed. Any one individual question may not fully cover all aspects of the LT as it will be assessed; be sure to review the example problems in your notes as well. 

***


**For this lab, you should begin by reading the accompanying primer and referring to your notes from in-class tutorials and in-class activities.**



Run this code chunk before you start each session. You do not yet need to understand it.
```{r}
#| label: setup
#| include: false

library(tidyverse)
library(mosaic)
library(ggformula)
```



## Part 1: Data Organization 

<span style='color:#000000;background:#D8BFD8;'>
<b>Section checked for completeness.</b></span>

### 1.1. 

Is this block of raw data organized correctly? Why?  
                     
Length (mm)   | Sex                 
------------- | -------------        
145 mm        | F                    
163 mm        | F                     
132 mm        | M                    
143 mm        | M                    

> No; you can't include units in the entered data. It can only be the numbers to be analyzed correctly.



### 1.2. 

Is this block of raw data organized correctly? Why? 

Info         | A       | B       | C       | D     
------------ | ------- | ------- | ------- | -------  
Length       | 145     | 163     |  132    |  143     
Sex          | F       | F       |  M      |  M  

> No; Rows are always supposed to be the individuals and columns are the measured variables. This block of data is backwards.  



***

## Part 2: Steph Curry

<span style='color:#000000;background:#D8BFD8;'>
<b>Section checked for completeness.</b></span>

Stephen Curry is one of the most prolific scorers currently in the NBA. We are interested in the number of points he scores across his entire career, from college to present day. The data of the number of points Stephen Curry scored during each regular season game from 2024-2025 can be found in `curry2024.csv`.  


### 2.1. (LT 1)  

Identify the population and the sample.  

a. Stephen Curry  
b. all NBA games  
c. all NBA games in the 2024-2025 season  
d. all games played by Curry  
e. all games played by Curry in the 2024-2025 season 
f. all points scored by Curry  

> The population is ... d .  
> The sample is ... e.  




### 2.2. (LT 2)  

Identify the variable type for *number of points scored by Curry*.  

a. qualitative nominal  
b. qualitative ordinal  
c. **quantitative discrete**  
d. quantitative continuous   

> c.



### 2.3. 

Read in the csv file `curry2024.csv` and store it as the object `curry` by modifying the code below to fill in each blank. Some blanks in this lab will have hints to the code you need written in the blanks like this: ___hint___ (that entire phrase, including underscores, is replaced). After reading in your data, it is a good habit to always use the function `names()` on your dataframe object to see the names of the variables.

```{r}
#| label: import-curry-data

curry <- read_csv("curry2024.csv", show_col_types = FALSE) # read.csv() is also ok
names(curry) # will print out the variable names
```



### 2.4. (LT R1)  

Find the mean number of points Curry scored in the 2024-2025 season by modifying the code below to replace the blanks with the correct code. If your answer looks strange but your code didn't produce an error, continue to the next question.

```{r}
#| label: calculate-mean-score

mean(~ Point, data = curry)
```

If you are confused, keep working through 2.6...



### 2.5. (LT R1)  

Insert a code chunk and write the code to print your raw data as the output. (Hint: refer to the Lab 1 Primer.) Why do you think you got the answer you did in 2.4? Hint: You may need to scroll through a couple rows of the data.

**Inserted code chunk**
```{r}
#| label: answer-2.5

curry
```


> There are some games that Curry didn't play, so he didn't score any points, and there are NA's in the dataset. R can't calculate a mean if there are 'words' in the dataset. (Luckily, NA's are special words that we can tell R to ignore and calculate the mean points scored in games Curry did play!)



### 2.6. (LT R1)  

We need to add an argument to `mean()` to calculate the mean number of points in games Curry played. The games he didn't play are coded with an `NA`. We need to tell R to ignore them for the sake of the `mean()` function. The argument we need to use to do that is called `na.rm`. Modify the code below to name the second argument. What is the mean number of points Curry scores per game in which he plays?  

```{r}
#| label: calcuate-mean-score-na

mean(~Point, data = curry, na.rm = TRUE)
```

> Curry averages `r round(mean(curry$Point, na.rm=TRUE), 4)` points per game in the 2024-2025 season.


### 2.7. (LT 2)  

Is the quantity you calculated in the previous question a parameter or a statistic? Explain your answer.  

> Statistic, as it was calculated on the sample of games Curry played in the 2024-2025 season [and not the population of all games he played].  


### 2.8. (LT R1)

Correct the following codes so that they do not return errors.  

```{r}
#| label: error-1

#sum(~ point, data = curry)
sum(~ Point, data = curry) 
# or 
sum(~ Point, data = curry, na.rm = TRUE)
```

> There was no variable `point`. R is case-sensitive, so it can't take a mean of a variable that doesn't exist.


```{r}
#| label: error-2

#length(curry$Game
length(curry$Game)
```

> The end parentheses was missing. R doesn't know the command is over until the end parentheses.


```{r}
#| label: error-3

#mean(~ Point, data = curry na.rm = TRUE)
mean(~ Point, data = curry, na.rm = TRUE)
```

> The comma was missing between arguments. 



***

## Part 3: Annual Greenhouse Gas Emissions    

<span style='color:#000000;background:#D8BFD8;'>
<b>Feedback given for this section.</b></span>

Greenhouse gas, mostly composed of carbon dioxide (CO~2~), is the major contributor of global warming and climate change. Most greenhouse gases are directly produced by the consumption of fossil fuel for energy, such as in transportation. Mauna Loa Observatory first began recording atmospheric carbon dioxide concentrations in 1958, but it wasn't until the 1970s that scientific concern and monitoring broadened. We are interested in exploring countries' annual greenhouse emissions in the years since 1970.

Researchers monitor gas emissions and record countries' annual greenhouse gas emissions yearly in kilotons of CO~2~ / 100,000. The data of the annual greenhouse emissions seen in Brazil, India, and the United States for each year from 1970 to 2012 are found in `greenhouse.csv`.



### 3.1. (LT 1)

What is the population and sample for this scenario?  

> Population: years since scientific concern began / years since 1970 / 1970s-current      
> Sample:  1970-2012       
  

<font size = 2>
<p style='color:#000000;background:#fcd283;'>
_**5 points**_
**2.5 points: population**   
......2.5 points if all years since 2007 (or alternative above)             
......2 points if missing [parts] (no [parts] this scenario)   
......1.5 points if have population, but extra information (i.e., variable)   
......0.5 points if wrote sample as the population        
......0 points if not about years     
**2.5 points: sample**   
......2.5 points if indicate 1970-2012       
......2 points if missing [parts]? (no [parts] this scenario)   
......1.5 points if have sample, but extra information (i.e., variable)  
......1 points if they have a reasonable selection, of their identified population, but not the correct sample   
......0.5 point if they have something about years, but the sample and population are not aligned    
......0 points if not about years      
...max 3 points if combined variable with population and/or sample 
in point tracker, 0 pts = answered and earned 0 points. NA = not answered.  
</p></font> 


<span style='color:#000000;background:#fdffcc;'>
<b>Potential Feedback</b>  
- Be more specific in how your sample is a subset of the population  
- This is combining the population and the variable being measured on that population.  
- What is one case in this dataset? The population is then all cases, and the sample is the selected cases.  
- Look for the population inside the research question.  
- You must indicate how the sample is a subset of the population, you identified different cases in the sample and population.  
</span>



### 3.2. 

Modify the code chunk below to read in your data and store it as `greenhouse`. Then modify the second line to print out *just* the variable names. 

```{r}
#| label: read-greenhouse-data

greenhouse <- read_csv("greenhouse.csv", show_col_types = FALSE)  #read.csv() is also fine
names(greenhouse)
```



<span style='color:#000000;background:#fdffcc;'>
<b>Potential Feedback</b>  
- read.csv is ok, but read_csv provides more options for managing data.   
- There is a code you learned that will display *only* the variable names.  
- You should use the code taught in class, not from the internet/friends/etc.  
- Almost, but you have a [spelling/capitalization] mistake that is keeping the code from running.  
- Almost, but you need the file extension (.csv) when you load in a file.  
</span>



### 3.3. (LT R1)  

What is the mean number of greenhouse gas emitted from 1970 t0 2012 by the United States?? Insert a code chunk and write the appropriate code to answer the question. Recall, you can enter a code chunk by using the following shortcuts: for Windows OS press Ctrl+Alt+I; for Mac OS Cmd+Opt+I.  

**Inserted code chunk**
```{r}
#| label: answer-3.3
# they do not need the label to get credit
mean(~ united_states, data = greenhouse)
```


<span style='color:#000000;background:#fdffcc;'>
<b>Potential Feedback</b>  
- Please use the formula template as taught in class in the future.  
- Same comment re: formula template as above.  
- One unit = one case; you can't find a mean of unit, just across units.  
- Taking the mean of the dataframe returns the mean value across the entire dataframe, not of the variable desired.  
- Code must occur in code chunks to run and show the output in your knitted document.  </span>



### 3.4. (LT R1)  

Which country, Brazil or India, had the most greenhouse gas emissions seen in a single year? Use the function `max()` (uses the formula template) to answer this question. (Hint: you should be using more than one line of code here.)        

**Inserted code chunk**
```{r}
#| label: answer-3.4
max(~ brazil, data = greenhouse) > max(~ india, data = greenhouse) # probably the the two lines below since I didn't teach logical operators
max(~ brazil, data = greenhouse) 
max(~ india, data = greenhouse)
```

> India    


<span style='color:#000000;background:#fdffcc;'>
<b>Potential Feedback</b>  
- Please use the formula template as taught in class in the future.  
- Same comment re: formula template as above.  
- One year = one case; you can't find a max of year, just across years.  
- Taking the max of the dataframe returns the max value across the entire dataframe, not of the variable desired.  
- To answer this question, you need to calculate both maximums to be able to determine which is larger.  
- You calculated the range, not the maximum.  
- `max()` can only be run on one variable at a time.  
- if multiple variables in one max function: Come talk to an instructor to discuss what this code is doing and how we can find the correct answer.  </span>


### 3.5. (LT R1)  

How much greenhouse gas in kilotons/100,000 were emitted in India between 1970 - 2012?  

**Inserted code chunk here**
```{r}
#| label: answer-3.5

sum( ~ india, data = greenhouse)
```

>   658.6083 kilotons/100,000 greenhouse emissions by India from 1970 to 2012. 


<span style='color:#000000;background:#fdffcc;'>
<b> Potential Feedback:</b>  
- `sum()` actually uses the formula template, so you can use it here too.  
- Same comment re: formula template as above.  
- Remember to include context / units in your written answers in the future.  
</span>



 <!-- GRADE ALL CODE Q's 3.2-3.5 TOGETHER -->
<p style='color:#000000;background:#fcd283;'>
**5 points: code**   
....5 points if fully correct use of functions with formula template     
....4 points if functional and calculates correct value, but not as taught (e.g.: not formula ~)   
....3 points if small mistake in code that prevent from running (spelling, capitalization, forgetting ~, etc.)    
....2 points if incorrect value calculated, but something done correctly  
....1 points if don't use functions taught in this class  
</p>

  


### 3.6. (LT R1)  

Researchers are interested in the total amount of greenhouse gas emitted for each year across the three countries. Add across vectors within each element and store the new vector as the object `total`. Print `total` to screen to see the total greenhouse gases in the combined countries each year. (Reminder: If you don't recall how to do something, you should refer to the primer from Lab #1.)  

**Inserted code chunk**
```{r}
#| label: answer-3.6
total<-greenhouse$brazil + greenhouse$india + greenhouse$united_states
total
```



<p style='color:#000000;background:#fcd283;'>
_**5 POINTS:**_  
**3 points: addition**   
....3 points if perfect: uses + to add vectors and stores   
....2 points if just missing a few columns/variables    
....1 points if small mistakes in code that prevent from running (spelling, capitalization, etc.)    
....1 points if uses sum() instead of +    
....1 points if uses sum() in addition to + (e.g. sum(variable + variable))    
....1 points if don't store       
**2 points: printing**    
...2 points if prints to screen using vector name only    
...1 point if uses print() to print to screen [not taught in this class]     
...0 points if missing  
**If they use code outside of class**     
...if they use rowSum or other code, or it looks AI generated include Warning below and they can get a 2 point max.    
...1 for assigning it (total <- )   
...1 if they "print" it correctly (just have total)
</p>

<span style='color:#000000;background:#fdffcc;'>
<b>Potential Feedback</b>  
- sum() adds across cases. You wanted to add across *variables* for each case, resulting in a vector of the total values across each unit.  Refer to your primer for a demonstration.   
- Refer to your primer for a demonstration of how to print to screen (as taught).   
- The c() function is unnecessary when using vectorized operators, like +   
- These calculations should be done and shown in the code    
- Needed to store and/or print to screen. Refer to your primer for a demonstration.    
- print() or rowSum() is not taught in this class -- please use the code taught in class and refer to the materials we give you; how to do this was taught in the primer. In future, we only give credit to code from the course.    
- You do not need to do this manually; refer to the Primer or see instructor for help.   
- = is not used for assignment.   
- Please use the code taught in class and refer to the materials we give you; how to do this was taught in the primer.   
</span>

<span style='color:#ffffff;background:#8b0000;'>
Warning: Code generated by ChatGPT or found on the internet is dangerous to use if you do not understand what it is doing. The course covers all code needed. All code needed was provided in the Lab 1 Primer. You may revise and resubmit using the code from the primer. In the future, code not used in class will not be graded.
</span>




***

## Part 4: Project Feederwatch 

<span style='color:#000000;background:#D8BFD8;'>
<b>Feedback given on this section</b></span>

Project FeederWatch, operated by The Cornell School of Ornithology and Birds Canada, collects citizen science data on feeder bird distribution and abundance. The unique bird monitoring tool is used to detect and explain the gradual changes in the wintering ranges of more than 100 species in North America.

Volunteers select a site that they consistently observe throughout the season from November to April. They pick multiple count periods, 48-hour window of time, to watch the site and keep a tally of the species that visit the site. Volunteers watch as often as they can and record how long and when they watch within each count period. At the end of the count period, the largest number of individuals per species seen simultaneously becomes the final tally that is reported to FeederWatch.

Your friend volunteers to watch their backyard the first weekend of every month of the 2021-2022 count season. Their filled tally sheets are in the file `FeederWatch_172885.pdf`. 


### 4.1. 

Create a spreadsheet (.csv) of the data from the field sheets in the file "FeederWatch_172885.pdf". *Think -- what is my individual unit? What were the characteristics measured on each unit? Now organize that according to the tidy data structures you learned*

You must include relevant information on the bird count summary, as well as the associated metadata for the nest. Metadata (literally 'beyond data') is the other relevant information associated with the variables measured on your cases. Part of the assignment is determining what should be included.  

You can enter the data in Excel (or other spreadsheet program), but be sure to save your file as a .csv (comma separated value; do not save it as a CSV UTF-8 file, a Macintosh csv or a MS dos csv--you want the regular CSV file called "Comma Separated Value (.csv)"). *(Note: you cannot have spaces in your vector/variable names; periods and underscores are the only symbols allowed.)* 

Name the file according to the file naming convention, with the suffix "_feeder-data.csv" (So Monte Rey would name his .csv file "ReyM_feeder-data.csv".) Upload the file to your project on the server. You will need to select this file to export along with your .qmd, .html, and .rproj files to turn in for this lab.





<p style='color:#000000;background:#fcd283;'>
_**5 POINTS:**_  
**1 point: csv** 
....1 point if csv file included, that has attempt at correct data entered  
....0 points if not csv or missing  
**2 points: tidy**  
....2 points if has bird count summary (blue columns) and rows organized as cases (correct = each date; you should not see something that matches the raw datasheets (included in .zip))    
....1.5 points if variable/column names have spaces   
<!-- ....1 point if they split a categorical variable into variables per level, but data otherwise tidily organized (meaning still rows as cases, so could make columns for species, but then size for one and NA for others in each row.) -->  
....1 point if tidy data but missing at least one blue variable    
....1 point if appears like it was attempted to be tidy    
<!-- (i.e.: close, but have a variable of "total" for all species)   -->
....0 points if missing all blue variables     
....0 points if not tidy    
**2 points: metadata**  
....2 points if two from yellow columns (Month, ID, Date, etc.) plus anything else included (yellow or green)    
....1 point if only one from yellow or green      
....1 point if missing all of yellow columns    
....0 points if missing   
</p>




<span style='color:#000000;background:#fdffcc;'>
<b>Potential Feedback</b>  
- What is a case in this instance? What is the unit that all variables are measured on? That should be your row.  
- Missing other information gathered.  
- Missing at least species and number  
- Variable names shouldn't include spaces, nor any symbols other than _  
- Okay, but this is not actually organized by case. You recorded data measured on each visit, not nest.   
- Okay, but this actually contains some information that is combined across cases, so your rows are no longer cases. See instructor to talk through scenario and understand the issue clearly.  
- Some 'data' you entered was not measured on a case, and was just part of organizing the fieldsheet.  
</span>




Now check to see that your file renders by clicking the button 'Render' at the top of this pane. If you have an error in your code, the file will not render and will show you where the error is occurring so you can fix it. Rendering to html will appear in the 'Viewer' tab in your "Files" pane if you click the gear tab and select "Preview in Viewer Pane." Recall that you MUST turn in a rendered .html with your lab; see the General Lab Instructions or the R Orientation Video Tutorials on rendering if you have trouble. 
